Transcriptomic responses to diet quality and viral infection in Apis mellifera

### 2019-06-09 Manuscript/Text/manuscript.tex
Accepted changes

For our nutrition factor, we examined two monofloral pollen diets, rockrose (\textit{Cistus} sp.) and chestnut (\textit{Castanea} sp.). Rockrose pollen is generally considered less nutritious than chestnut pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron \cite{adamInt, DiPasquale}. For our virus factor, one level contained bees that were inoculated with IAPV and another level contained bees that were not inoculated with IAPV. This experimental design resulted in four treatment groups (low quality rockrose pollen without IAPV exposure, high quality chestnut pollen without IAPV exposure, low quality rockrose pollen with IAPV exposure, and high quality chestnut pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honey bees.

@@

We observed a substantially larger number of differentially expressed genes (DEGs) in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1). There were only four genes that were DEGs in both our diet main effect and our virus main effect (GB48747, GB47214, GB42908, and GB42507). In the diet factor, more DEGs were upregulated in the more-nutritious chestnut group (\textit{n} = 1,033) than in the less-nutritious rockrose group (\textit{n} = 881). In the virus factor, there were more virus-upregulated DEGs (\textit{n} = 38) than control-upregulated DEGs (\textit{n} = 5). While these reported DEG counts are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Supplementary table 1, Additional file 1 and Additional file 18). For our DEG analysis, we used R software version 3.3.3 \cite{R}.

@@

The R-squared values between gene read counts and pathogen response measurements were generally low (R-squared $<$ 0.1) across our dataset (Supplementary table 9, Additional file 1). We further explored whether clusters of DEGs showed higher correlations with pathogen response measurements than non-DEGs (the latter serving as a control, where we do not expect a correlation). A Kruskal–Wallis test was used to determine if R-squared distributions of DEG clusters significantly differed from those in the rest of the data. The p-values and Bonferroni correction values for each of the 36 tests (as described in the methods section) is provided in Supplementary table 9, Additional file 1. An overall trend emerges to suggest that DEGs may have significantly larger correlation with the pathogen response measurements compared to non-DEGs.

@@

There were several limitations in this study that could be improved upon in future studies. For instance, our comparison between the Galbraith data (single-drone colonies) and our data (naturally-mated colonies) was limited by numerous extraneous variables between these studies. In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on worker honey bees that were fed sugar and artificial pollen diets, whereas our study focused on worker honey bees that were fed bee-collected monofloral diets. Also, Galbraith’s bees were sampled at 24 hours while ours were sampled at 36 hours. Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and observations to determine behaviorally symptomatic bees whereas we used whole bodies and categorized only into inoculated vs. non-inoculated groups. There are also differences in the hours post inoculation and possible differences in the inoculation amount between the studies. Further differences between the studies can be found in their corresponding published methods sections \cite{adamInt, galbraith}. The different factors between these two studies may be critical because particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed. This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, SBV vs. DWV \cite{Ryabov}. Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}.

@@

@@

\section*{Funding}
  This work was supported by the United States Department of Agriculture, Agriculture and Food Research Initiative (USDA-AFRI) 2011-04894.

\section*{Author's contributions}
  LR performed the bioinformatic and statistical analyses and produced the figures and tables. LR and ALT drafted the manuscript. ALT, AGD, JCT, BCB contributed to experimental design. AGD and JCT carried out laboratory experiments. AGD processed samples for virus and RNA-seq. DC advised on statistical analyses and visualization. All authors revised and approved the manuscript.

@@

  \subsection*{Additional file 1 --- Supplementary tables.}
    \textbf{Table 1:} Number of DEGs across three analysis pipelines for (A) the diet main effect in our study, (B) the virus main effect in our study, and (C) the virus main effect in the Galbraith study. For the diet effects, ``C'' represents chestnut diet and ``R'' represents rockrose diet. For the virus effects, ``N'' represents control non-inoculated and ``V'' represents virus-inoculated.
    \textbf{Table 2:} Pathways related to the 1,033 DEGs that were upregulated in the chestnut treatment from the diet main effect.
    \textbf{Table 3:} Pathways related to the 881 DEGs that were upregulated in the rockrose treatment from the diet main effect.
    \textbf{Table 4:} GO analysis results for the 601 DEGs that were upregulated in the NC treatment from the NC versus NR treatment pair analysis. These DEGs represent genes that are upregulated when non-inoculated honey bees are given high quality chestnut pollen compared to being given low quality rockrose pollen.
    \textbf{Table 5:} GO analysis results for the 340 DEGs that were upregulated in the NR treatment from the NC versus NR treatment pair analysis. These DEGs represent genes that are upregulated when non-inoculated honey bees are given low quality rockrose pollen compared to being given high quality chestnut pollen.
    \textbf{Table 6:} GO analysis results for the 247 DEGs that were upregulated in the VC treatment from the VC versus VR treatment pair analysis. These DEGs represent genes that are upregulated when inoculated honey bees are given high quality chestnut pollen compared to being given low quality rockrose pollen.
    \textbf{Table 7:} GO analysis results for the 129 DEGs that were upregulated in the VR treatment from the VC versus VR treatment pair analysis. These DEGs represent genes that are upregulated when inoculated honey bees are given low quality rockrose pollen compared to being given high quality chestnut pollen.
    \textbf{Table 8:} Number of DEGs across three analysis pipelines for all six treatment pair combinations between the diet and virus factor. ``C'' represents chestnut diet, ``R'' represents rockrose diet, ``V'' represents virus-inoculated, and ``N'' represents control non-inoculated.
    \textbf{Table 9:} Kruskal-Wallis p-value and Bonferroni corrections for the 36 combinations of DEG lists, pathogen response metrics, and cluster number. (XLS).

>>>>

For our nutrition factor, we examined two monofloral pollen diets, rockrose (\textit{Cistus} sp.) and chestnut (\textit{Castanea} sp.). Rockrose pollen is generally considered less nutritious than chestnut pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron \cite{adamInt, DiPasquale}. For our virus factor, one level contained bees that were inoculated with IAPV and another level contained bees that were not inoculated with IAPV. This experimental design resulted in four treatment groups (NR: low quality rockrose pollen without IAPV exposure; NC: high quality chestnut pollen without IAPV exposure; VR: low quality rockrose pollen with IAPV exposure; VC: high quality chestnut pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honey bees.

@@

In bees collected 36 hours post treatment, we observed a substantially larger number of differentially expressed genes (DEGs) in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1). There were only four genes that were DEGs in both our diet main effect and our virus main effect (GB48747, GB47214, GB42908, and GB42507). In the diet factor, more DEGs were upregulated in the more-nutritious chestnut group (\textit{n} = 1,033) than in the less-nutritious rockrose group (\textit{n} = 881). In the virus factor, there were more virus-upregulated DEGs (\textit{n} = 38) than control-upregulated DEGs (\textit{n} = 5). While these reported DEG counts are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Supplementary table 1, Additional file 1 and Additional file 18). For our DEG analysis, we used R software version 3.3.3 \cite{R}.

@@

The R-squared values between gene read counts and pathogen response measurements were generally low (R-squared $<$ 0.1) across our dataset (Supplementary table 9, Additional file 1). We further explored whether clusters of DEGs showed higher correlations with pathogen response measurements than non-DEGs (the latter serving as a control, where we do not expect a correlation). A Kruskal–Wallis test was used to determine if R-squared distributions of DEG clusters significantly differed from those in the rest of the data. The p-values and Bonferroni correction values for each of the 36 tests (as described in the methods section) is provided in Supplementary table 9, Additional file 1. Distribution of the R-squared values for DEG cluster read counts and pathogen response metrics is provided in Additional file 19. An overall trend emerges to suggest that DEGs may have significantly larger correlation with the pathogen response measurements compared to non-DEGs.

@@

There were several limitations in this study that could be improved upon in future studies. For instance, our comparison between the Galbraith data (single-drone colonies) and our data (naturally-mated colonies) was limited by numerous extraneous variables between these studies.

In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on worker honey bees that were fed sugar and artificial pollen diets, whereas we used whole bodies and categorized only into inoculated vs. non-inoculated groups; noise may have been introduced through different responses in asymptomatic bees. Also, Galbraith’s bees were sampled at 24 hours while ours were sampled at 36 hours. Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and observations to determine behaviorally symptomatic bees whereas we used whole bodies and categorized only into inoculated vs. non-inoculated groups. There are also differences in the hours post inoculation and possible differences in the inoculation amount between the studies. Further differences between the studies can be found in their corresponding published methods sections \cite{adamInt, galbraith}. The different factors between these two studies may be critical because particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed. This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, SBV vs. DWV \cite{Ryabov}. Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}.

@@

\section*{Abbreviations}

IAPV: Israeli acute paralysis virus;  DWV: deformed wing virus; SBV: sacbrood virus; NR: non-inoculation, low quality (\textit{Cistus} sp.) pollen treatment; NC: non-inoculation, high quality pollen (\textit{Castanea} sp.) pollen treatment; VR: IAPV-inoculated, low quality (\textit{Cistus} sp.) pollen treatment; VC: IAPV-inoculated, high quality pollen (\textit{Castanea} sp.) pollen treatment; REML: restricted maximum likelihood;
DEG: differentially expressed gene; GO: gene ontology; PCD: principal component analysis.

@@

\section*{Funding}
  This work was supported by the United States Department of Agriculture, Agriculture and Food Research Initiative (USDA-AFRI) 2011-04894, but it had no role in the design, analysis, or interpretation of the study.

\section*{Author's contributions}
  LR performed the bioinformatic and statistical analyses and produced the figures and tables. LR, AGD, and ALT drafted the manuscript. All authors contributed to the conception of the project, the experimental design, and the interpretation of the data. AGD and JCT carried out laboratory experiments. AGD processed samples for virus and RNA-seq. DC advised on statistical analyses and visualization. All authors revised and approved the manuscript.

@@

  \subsection*{Additional file 1 --- Supplementary tables.}
  In all tables, ``C'' represents chestnut diet, ``R'' represents rockrose diet, ``N'' represents control non-inoculated, and ``V'' represents virus-inoculated.
    \textbf{Table 1:} Number of DEGs across three analysis pipelines for the (A) diet main effect in our study, (B) virus main effect in our study, and (C) virus main effect in the Galbraith study.
    \textbf{Table 2:} Pathways related to the 1,033 DEGs upregulated in the chestnut treatment from the diet main effect.
    \textbf{Table 3:} Pathways related to the 881 DEGs upregulated in the rockrose treatment from the diet main effect.
    \textbf{Table 4:} GO analysis results for the 601 DEGs upregulated in the NC treatment from the NC versus NR treatment pair analysis. These DEGs represent genes upregulated in non-inoculated honey bees given high quality chestnut pollen versus low quality rockrose pollen.
    \textbf{Table 5:} GO analysis results for the 340 DEGs upregulated in the NR treatment from the NC versus NR treatment pair analysis. These DEGs represent genes upregulated in non-inoculated honey bees given low quality rockrose pollen versus high quality chestnut pollen.
    \textbf{Table 6:} GO analysis results for the 247 DEGs upregulated in the VC treatment from the VC versus VR treatment pair analysis. These DEGs represent genes upregulated in inoculated honey bees given high quality chestnut pollen versus low quality rockrose pollen.
    \textbf{Table 7:} GO analysis results for the 129 DEGs upregulated in the VR treatment from the VC versus VR treatment pair analysis. These DEGs represent genes upregulated in inoculated honey bees given low quality rockrose pollen versus high quality chestnut pollen.
    \textbf{Table 8:} Number of DEGs across three analysis pipelines for all six treatment pair combinations between the diet and virus factor.
    \textbf{Table 9:} Kruskal-Wallis p-value and Bonferroni corrections for the 36 combinations of DEG lists, pathogen response metrics, and cluster number. (XLS).
